Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-11 _tex/paper2.tex
Built site for gh-pages

Figure~\ref{fig-cj04} illustrates the preliminary integration of
theoretical principles into the conceptual-population model.
Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
motivating example. This SCM maps the relationship between the
conceptual-population outcome \((O^{cp}_{abhijk})\) to several other
variables. In this SCM, the subscripts \(\{a,b\}\) represent the
compared texts (stimuli), with \(a \neq b\). The indices \(\{h,i\}\)
label the students who wrote the texts (individuals), where \(h = i\) is
possible. Moreover, the index \(j\) denotes the judge, and \(k\) denotes
the judgment number, which accounts for experimental cases where judges
perform multiple judgments of the same texts written by the same
students, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}.

The other variables in SCM~\ref{fig-cj04_scm} include the texts'
discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
difference \((D_{abhijk})\), in line with Thurstone's theory (see
Section~\ref{sec-thurstone_theory}). Additionally, based on the evidence
presented in Section~\ref{sec-theory-issue1b} and the recommendations of
\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, we incorporate the
judges' biases \((B_{kj})\) into the model. Together with the outcome,
these variables form the preliminary set of endogenous variables,
\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
Finally, the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

By incorporating judges' biases into the conceptual-population model, we
assume that additional relevant characteristics of the stimuli are
unlikely to be equally weighted or opposing. Thus, these characteristics
may exert an uneven influence on judges' perceptions, creating biases in
their judgments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases may ultimately introduce dependencies between
the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}, hence the need
for their integration. The full statistical elaboration of this point is
presented in Section~\ref{sec-statistical}.

@@

\[
\begin{aligned}
  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
\end{aligned}
\]

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering the advantage of
highlighting potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}. Thus, this study uses the magnified
representation.

Figure~\ref{fig-cj04_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ experiments. It also distinguishes between observable
endogenous variables, such as the outcome, and variables considered
latent according to the theory and practice of CJ, such as the the
text's discriminal processes, their discriminal difference, and the
judges' biases.

It is easy to see that representing the CJ system using indices for a
single pair of individuals and stimuli, along with the indices for the
judgment number and the judge responsible for the comparison, can
quickly become intractable and cumbersome. Thus, to maintain generality
and simplify the notation, certain indices are grouped into vectors or
matrices. First, we define the matrices \(IA\) and \(JK\) as:

\[
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
\]

Each row of \(IA\) represents a unique pairing of an individual
\(\{h,i\}\) with a stimulus \(\{a,b\}\). Here, \(n_{I}\) represents the
number of individuals and \(n_{A}\) the number of stimulus within
individuals in the population. Similarly, each row of \(JK\) represents
a unique pairing of a judge \(j\) with a judgment \(k\). Here, \(n_{J}\)
represents the number of judges and \(n_{K}\) the number of judgments
performed by each judge in the population.

@@

\subsection{Statistics and Causal inference}\label{sec-appendixB}

>>>>

SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
theoretical principles into the conceptual-population model. This SCM
maps the relationship between the conceptual-population outcome,
\((O^{cp}_{abhijk})\), and several related variables. The subscripts
\(a\) and \(b\) represent the compared texts (stimuli), while the
indices \(h\) and \(i\) label the students who authored them
(individuals). Moreover, the index \(j\) refers to the judge, and \(k\)
specifies the judgment number, accounting for experimental conditions
where judges compare the same texts written by the same students
multiple times, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}. It is important to note that the
indexing system allows for comparisons between different texts written
by the same student (\(h = i\), \(a \neq b\)) and between texts written
by distinct students (\(h \neq i\), where \(a = b\) is possible).
However, it excludes cases where judges compare a student's text to
itself (\(h = i\), \(a = b\)), as such a comparison lacks practical
meaning.

The other variables in SCM~\ref{fig-cj03_scm} include the texts'
discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
difference \((D_{abhijk})\), in line with Thurstone's theory (see
Section~\ref{sec-thurstone_theory}). Additionally, based on the
arguments presented in Section~\ref{sec-theory-issue1b} and the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
the SCM incorporates the judges' biases \((B_{kj})\). Together with the
outcome, these variables form the preliminary set of endogenous
variables,
\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
Finally, the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\[
\begin{aligned}
  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
\]

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another showing the direction of causal
influence. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs depict observed variables as solid black
circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. The \emph{standard representation} of DAGs
typically omits exogenous variables for simplicity, while the
\emph{magnified representation} includes them to highlight potential
issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}. This section adopts the magnified
representation to enhance clarity.

Figure~\ref{fig-cj03_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ's experiments. It also distinguishes between observable
endogenous variables, such as the outcome, and variables considered
latent according to the theory and practice of CJ, such as the text's
discriminal processes, their discriminal difference, and the judges'
biases.

Although SCMs and DAGs do not necessitate specifying the data structure
of the modeled system, defining these structures can sometimes enhance
their clarity and simplify their notation. Thus, we organize certain
indices into vectors or matrices. First, we define the matrices \(IA\)
and \(JK\) as follows:

\begin{equation}\phantomsection\label{eq-mat12}{
IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N} \; ; \quad
JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
}\end{equation}

where each row of \(IA\) corresponds to a unique pairing of individuals
\(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents the number of
individuals, while \(n_{A}\) denotes the number of stimuli available per
individual. Thus, the matrix \(IA\) is composed of natural numbers
\(\mathbb{N}\), with \(n_{I} \cdot n_{A}\) rows and \(2\) columns.
Similarly, each row of \(JK\) specifies a unique pairing of a judge
\(j\) with a judgment \(k\). In this case, \(n_{J}\) indicates the total
number of judges, and \(n_{K}\) represents the number of judgments each
judge performs. Therefore, the matrix \(JK\) is also composed of natural
numbers \(\mathbb{N},\) with \(n_{J} \cdot n_{K}\) rows and \(2\)
columns.

Importantly, since the discriminal difference of the stimuli becomes an
observable outcome only through judges' perceptions, it is necessary to
define the matrix \(RPUV\), where each row of the matrix uniquely pairs
the corresponding rows from the matrices \(IA\) and \(JK\) as follows:

\begin{equation}\phantomsection\label{eq-mat3}{
RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & a & i & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
}\end{equation}

Here, the \(RPUV\) matrix is also composed of natural numbers
\(\mathbb{N}\), and has a dimension of
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
columns. The dimension of the matrix results from the
conceptual-population model's assumption that researchers have access to
data representing all repeated comparisons made by every available judge
for each pair of stimuli produced by each pair of individuals in the
population. Here, the term \({n_{I} \cdot n_{a} \choose 2}\) represents
the Binomial coefficient, which quantifies the number of
individual-stimuli comparison pairs, and it is defined as:

\[
{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
\]

Using the motivating example outlined in Section~\ref{sec-theoretical},
we can visualize the shape of these structures. For illustrative
purposes, assume the conceptual population consists of \(n_{I} = 5\)
individuals, each with \(n_{A} = 2\) available stimuli. Additionally,
assume that each pair of stimuli is evaluated by \(n_{J} = 3\) judges,
with each judge making \(n_{K} = 3\) comparative judgments. Given this
setup, the matrices in (\ref{eq-mat12}) and (\ref{eq-mat3}) are
represented as follows:

\[
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix}\; ; \quad
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} \; ; \quad
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

With these matrices we can express an equivalent version of
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a vectorized form.

@@

\subsection{Statistical and Causal inference}\label{sec-appendixB}
